---
title: "Variance Factorizations"
author: "Georges Monette <georges@yorku.ca>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r}
library(magrittr)

```

## Variance Factorizations

The Log-Cholesky factorization used for the variance-covariance matrix of random effects in `nlme` does not lend itself easily to selectively modeling some random effects as independent.  This stems from the fact that the factorization is a 'right' factorization in which the variance matrix is expressed as:
$$
V = R'R
$$
where $R$ is an upper-triangular (or 'right-triangular') matrix. 

Constraining off-diagonal upper-triangle elements of $R$ to be zero does not result in the corresponding element of $V$ being zero. That is, it does not result in a zero covariance. It produces, however, a zero conditional covariance.

**TO DO:** Demonstrate above.

On the other hand, a left-triangular factorization:
$$
V = L'L
$$
will have the property that 0 elements of $L$ in the lower triangle will result in 0 elements of $V$ in the corresponding position.

The `chol` function in returns a right factorization. Note that the use of the `pivot` argument to handle rank-deficient matrices produces a factor that is not triangular if a non-trivial pivot is applied.  As a result, pivotting cannot be used to handle singular variances since the assumption of triangular parametrization would be violated.

```{r}
(V <- diag(1:3) + 1)
(f <- chol(V))
t(f) %*% f
f[2,3] <- 0
f
(v2 <- t(f) %*% f)
# The conditional covariance is zero
v2[2:3,2:3] - 
  v2[2:3,1,drop=FALSE] %*% 
    solve(v2[1,1],drop=FALSE) %*%
    v2[1,2:3,drop=FALSE]
# How to do a left-factorization
cholR <- function(x, ...) {
  # this can return a non-triangular factor, thus can't  be used for our purposes
  ret <- chol(x, pivot = TRUE)
  pivot <- attr(ret, "pivot")
  ret[, order(pivot)]
  }

cholL <- function(x, ...) {
  x[,] <- rev(x)
  ret <- chol(x) # not cholR (see above)
  ret[,] <- rev(ret)
  ret
  }
# test rank deficient V
V0 <- crossprod( cbind(1, 1:10, (1:10)^2, 11:20))
tryCatch(chol(V0), error = function(e) e)
tryCatch(cholR(V0), error = function(e) e)


lfac <- cholL(V)  
t(lfac) %*% lfac
lfac[3,2] <- 0
t(lfac) %*% lfac
```

### Is there an efficient way to turn a right factor into a left factor?

An inefficient, and probably numerically undesirable, way would construct the variance matrix and perform a left factorization: `cholL( t(R) %*% R)`.

Instead, we use the `qr` decomposition of `R`, avoiding computing the variance matrix and the ensuing loss of precision.

```{r}
(lfac <- cholL(V))
rfac <- qr.R(qr(lfac))
chol(V)
t(rfac) %*% rfac - (t(lfac) %*% lfac)
X <- cbind(1,1:10,(1:10)^2,11:20)
qr.R(qr(X))
qr.R(qr(X, pivot = T))

L2R <- function(x, ...) {
    R <- qr.R(qr(x))
    sign(diag(R)) * R
  }
V

cholL(V) %>% round(2)
V - crossprod( cholL(V))
V - crossprod( chol(V))

V - crossprod( L2R(cholL(V)))


R2L <- function(x, ...) {
    x[] <- rev(x)
    ret <- L2R(x, ...)
    ret[] <- rev(ret)
    ret
  }
x <- matrix(rnorm(10000), 100)
system.time(
    for(i in 1:1000) R2L(x)
  )
system.time(
    for(i in 1:1000) L2R(x)
  )
# Check that it works

V
Vr <- crossprod( matrix(rnorm(9),3))
svd(Vr)$d
lc <- cholL(Vr)
Vr - crossprod(lc)
rc <- chol(Vr)
Vr - crossprod(rc)
lc2r <- L2R(lc)
Vr - crossprod(lc2r)
rc2l <- R2L(rc)
Vr - crossprod(rc2l)
```
## Creating a pdClass based on pdLogChol

Seems that:

- we need pdFactor.pdInd which is called by pdMatrix etc.
- we need solve.pdInd which returns the 'pdInd' object of the inverse (we can take advantage of a few things here)
- we might not need coef.pdInd since coef.pdMat would work, but perhaps might as well
- we need, of course, pdInd and, I'm pretty sure, pdConstruct.pdInd

Note the following:

- solve.pdInd does not construct an inverse, just the pdInd representation of the inverse.
- So, where is the inverse right-factor constructed -- which I assume is needed for the ML? Perhaps pdFactor, which is exported by nlme. 
- The C function `logChol_pd` in `pdFactor.pdLogChol` returns a list of three components:
1. The R factor as a vector, listed column by column
2. The number of rows in R, as a single number
3. The logChol coeficients: log of diagonal elements of R followed by off-diagonal elements
- A call to pdMatrix for a pdLogChol/pdSymm/pdMat object calls the 'pdMatrix.pdSymm' method that in turn calls 'pdFactor' if the argument 'factor' is TRUE. Otherwise, it calls pdMatrix.pdMat that calls `crossprod(pdMatrix(object,factor=TRUE))`
- So, bottom line, `pdFactor` is the critical method and we need a `pdFactor.pdInd`.  

## solve.pdInd

The solve method transforms a pdInd object representing a variance matrix to a pdInd object representing the inverse of that matrix.  If we let $L$ be a left-triangular factorization of $V = L'L$ then, if $F$ is the left-triangular factorization of $V^{-1}$, we have that $ F = R'^{-1}$ where $R$ is the right-triangular factor such that $V = R'R$.

Thus, `solve.pdInd` would transform $L$ to $R'^{-1}$.

To take advantage of the structure of a `pdInd` object, which is a factorization of a variance $V$ of the form:

\[V = \left[ {\begin{array}{*{20}{c}}
  A&{B'} \\ 
  B&C 
\end{array}} \right]\]

where $C$ is diagonal. In this case, $L$ has the form:

\[L = \left[ {\begin{array}{*{20}{c}}
  D&0 \\ 
  E&F 
\end{array}} \right]\]

conformably with the partitioning of $V$ with $F$ a diagonal matrix. 

Now an important problem, in contrast with existing methods for pdMat objects, is that the left-triangular factor of the inverse cannot be expressed with the unconstrained parametrization for $V$.  

If the result of `solve` is only used through `pdFactor` or `pdMatrix`, then it is a simple thing to set a flag for the `pdInd` object indicating that it represents the inverse. We can give `pdFactor` and `pdMatrix` the job of computing the correct right-triangular factor. 

If the object created by `solve.pdInd` is used in other ways then the problem might be more difficult.

One hopes that the approach of using `solve` to create the representation of $V^{-1}$ arose simply from the fact that the `pdMat` classes in the original version of `nlme` all lent themselves to this approach.

Here are some relevant expressions:

\[\begin{gathered}
  V = \left[ {\begin{array}{*{20}{c}}
  A&{B'} \\ 
  B&C 
\end{array}} \right] \hfill \\
  L = \left[ {\begin{array}{*{20}{c}}
  D&0 \\ 
  E&F 
\end{array}} \right] \hfill \\ 
\end{gathered} \]

\[\left[ {\begin{array}{*{20}{c}}
  A&{B'} \\ 
  B&C 
\end{array}} \right] = \left[ {\begin{array}{*{20}{c}}
  {D'}&{E'} \\ 
  0&F 
\end{array}} \right]\left[ {\begin{array}{*{20}{c}}
  D&0 \\ 
  E&F 
\end{array}} \right] = \left[ {\begin{array}{*{20}{c}}
  {D'D + E'E}&{E'F} \\ 
  {FE}&{{F^2}} 
\end{array}} \right]\]

\[F = {C^{1/2}}\]

\[E = {F^{ - 1}}E\]

\[D'D = A - E'E\]

\[R = \left[ {\begin{array}{*{20}{c}}
  G&H \\ 
  0&K 
\end{array}} \right]\]

\[V = R'R = \left[ {\begin{array}{*{20}{c}}
  A&{B'} \\ 
  B&C 
\end{array}} \right] = \left[ {\begin{array}{*{20}{c}}
  {G'}&0 \\ 
  {H'}&{K'} 
\end{array}} \right]\left[ {\begin{array}{*{20}{c}}
  G&H \\ 
  0&K 
\end{array}} \right] = \left[ {\begin{array}{*{20}{c}}
  {G'G}&{G'H} \\ 
  {H'G}&{H'H + K'K} 
\end{array}} \right]\]

The problem arises from the fact that, in the case of the right-triangular factorization, the diagonality of $C$ does not imply the diagonality of $K$, which is precisely why we need to use the left-triangular factorization in the first place.

Thus, when we invert and transpose $R$ to get the left-triangular factor of $V^{-1}$ we cannont use the parsimonious unrestricted parametrization of the `pdInd` object.

To get around this, we will attempt to use the following strategy:

1. Add an 'invert' attribute to 'pdInd' object. Intially, it is set to FALSE.
2. `solve.pdInd` simply inverts the flag (from FALSE to TRUE or TRUE to FALSE).
3. `pdFactor.pdInd` returns a right-triangular factor for $V$ or for $V^{-1}$ depending on the value of the 'invert' attribute.

Thus the burden of dealing with inversion is passed along to `pdFactor.pdInd`.








```{r}
library(nlme)
methods(class="pdLogChol")
methods(class="pdSymm")
nlme:::coef.pdSymm

methods(class="pdMat")

nlme:::pdLogChol
nlme:::pdSymm
nlme:::pdMat

pdFactorfull <- function (object) 
{
    Ncol <- round((-1 + sqrt(1 + 8 * length(object)))/2)
    .C(logChol_pd, Factor = double(Ncol * Ncol), as.integer(Ncol), 
        as.double(object))
}
environment(pdFactorfull) <- environment(pdFactor)



```


## Examples

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
